\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}

### 2021-07-09 Manuscripts/Manuscript by chapters - Wiley submission/00-Full_document_No_Appendix.tex
Created folder for Wiley submission

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Using generalized additive models to analyze biomedical non-linear longitudinal data},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\newcommand{\passthrough}[1]{#1}
\lstset{defaultdialect=[5.3]Lua}
\lstset{defaultdialect=[x86masm]Assembler}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{lineno}
\usepackage[affil-it,blocks]{authblk}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage[nomarkers,figuresonly]{endfloat}
%set a box to put the ORCID logo
\newbox{\myorcidaffilbox}
\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{orcid}}

@@

\usepackage{placeins}
\usepackage{subfig}
\usepackage{breqn}
\usepackage[font={small}]{caption}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\doublespacing

\hypertarget{summary}{%
\section{Summary}\label{summary}}

@@

cancer biology; tumor response; generalized additive models; simulation; R

@@

Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study). Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others. Tumor response (\protect\hyperlink{ref-roblyer2011}{Roblyer \emph{and others} 2011}; \protect\hyperlink{ref-demidov2018}{Demidov \emph{and others} 2018}; \protect\hyperlink{ref-pavlov2018}{Pavlov \emph{and others} 2018}; \protect\hyperlink{ref-tank2020}{Tank \emph{and others} 2020}), antibody expression (\protect\hyperlink{ref-ritter2001}{Ritter \emph{and others} 2001}; \protect\hyperlink{ref-roth2017}{Roth \emph{and others} 2017}), and cell metabolism (\protect\hyperlink{ref-skala2010}{Skala \emph{and others} 2010}; \protect\hyperlink{ref-jones2018}{Jones \emph{and others} 2018}) are examples of the different situations where researchers have used longitudinal designs to study some physiological response. Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals(\protect\hyperlink{ref-greening2018}{Greening \emph{and others} 2018}), to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients (\protect\hyperlink{ref-sio2016}{Sio \emph{and others} 2016}), to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients (\protect\hyperlink{ref-kamstra2015}{Kamstra \emph{and others} 2015}).

Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study. The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times (\protect\hyperlink{ref-wagenmakers2008}{Wagenmakers \emph{and others} 2008}), and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA). The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) (\protect\hyperlink{ref-gueorguieva2004}{Gueorguieva and Krystal 2004}; \protect\hyperlink{ref-schober2018}{Schober and Vetter 2018}).

The expected linear behavior of the response through time is a key requisite in rm-ANOVA (\protect\hyperlink{ref-pinheiro2006}{Pinheiro and Bates 2006}). This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference. In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies. A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings (\protect\hyperlink{ref-vishwanath2009}{Vishwanath \emph{and others} 2009}; \protect\hyperlink{ref-skala2010}{Skala \emph{and others} 2010}; \protect\hyperlink{ref-roblyer2011}{Roblyer \emph{and others} 2011}). These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation. Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.

A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data. In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs(\protect\hyperlink{ref-dennis2019}{Dennis \emph{and others} 2019}). For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid (\protect\hyperlink{ref-wang2019}{Wang \emph{and others} 2019}).

Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) (\protect\hyperlink{ref-liu2010}{Liu \emph{and others} 2010}; \protect\hyperlink{ref-halsey2015}{Halsey \emph{and others} 2015}), consequently biasing the conclusions of the study. Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni (\protect\hyperlink{ref-abdi2010}{Abdi 2010})), but they in turn reduce statistical power (1-\(\beta\))(\protect\hyperlink{ref-nakagawa2004}{Nakagawa 2004}), and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) (\protect\hyperlink{ref-gelman2012}{Gelman \emph{and others} 2012}; \protect\hyperlink{ref-albers2019}{Albers 2019}). Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.

On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases (\protect\hyperlink{ref-ugrinowitsch2004}{Ugrinowitsch \emph{and others} 2004}). Corrections can be made in rm-ANOVA in the absence of compound symmetry (\protect\hyperlink{ref-greenhouse1959}{Greenhouse and Geisser 1959}; \protect\hyperlink{ref-huynh1976}{Huynh and Feldt 1976}), but the effectiveness of the correction is limited by the size of the sample, the number of measurements(\protect\hyperlink{ref-haverkamp2017}{Haverkamp and Beauducel 2017}), and group sizes (\protect\hyperlink{ref-keselman2001}{Keselman \emph{and others} 2001}). In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons (\protect\hyperlink{ref-charan2013}{Charan and Kantharia 2013}) which might cause the corrections for lack of compound symmetry to be ineffective.

Due to a variety of causes, the number of observations during a study can vary between all subjects. For example, in a clinical trial patients may voluntarily withdraw, whereas attrition due to injury or weight loss in preclinical animal studies is possible. It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points. In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated. When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis (\protect\hyperlink{ref-gueorguieva2004}{Gueorguieva and Krystal 2004}). This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects. At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data. The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data (\protect\hyperlink{ref-vishwanath2009}{Vishwanath \emph{and others} 2009}; \protect\hyperlink{ref-skala2010}{Skala \emph{and others} 2010}). Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age). When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject (\protect\hyperlink{ref-pinheiro2006}{Pinheiro and Bates 2006}). However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent (\protect\hyperlink{ref-gueorguieva2004}{Gueorguieva and Krystal 2004}; \protect\hyperlink{ref-barr2013}{Barr \emph{and others} 2013}). And even more importantly, LMEMs also assume a linear relationship between the response and time (\protect\hyperlink{ref-pinheiro2006}{Pinheiro and Bates 2006}), making them unsuitable to analyze non-linear data.

As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference. In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data. Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data. Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data (\protect\hyperlink{ref-rose2012}{Rose \emph{and others} 2012}; \protect\hyperlink{ref-simpson2018}{Simpson 2018}; \protect\hyperlink{ref-pedersen2019}{Pedersen \emph{and others} 2019}), health-environment interactions (\protect\hyperlink{ref-yang2012}{Yang \emph{and others} 2012}) and the dynamics of government in political science (\protect\hyperlink{ref-beck1998}{Beck and Jackman 1998}). There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements (\protect\hyperlink{ref-wood2017}{Wood 2017}) and 3) can easily accommodate missing observations. Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.

The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs. In particular, \(\textsf{R}\) (\protect\hyperlink{ref-r}{R Core Team 2020}) has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} (\protect\hyperlink{ref-wood2016}{Wood \emph{and others} 2016}; \protect\hyperlink{ref-wood2017}{Wood 2017}) that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user. At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models (\protect\hyperlink{ref-haverkamp2017}{Haverkamp and Beauducel 2017}). Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.

This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas. First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations are explained in detail. Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations. Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies (\protect\hyperlink{ref-vishwanath2009}{Vishwanath \emph{and others} 2009}). The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies. Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in \(\textsf{R}\), in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.

@@

A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish (\protect\hyperlink{ref-pinheiro2006}{Pinheiro and Bates 2006}; \protect\hyperlink{ref-west2014}{West \emph{and others} 2014}). A LMEM with interaction between time and treatment for a longitudinal study can be written as:

@@

{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_No_Appendix_files/figure-latex/l-q-response-1}

@@

{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_No_Appendix_files/figure-latex/basis-plot-1}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Using generalized additive models to analyze biomedical non-linear longitudinal data},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\newcommand{\passthrough}[1]{#1}
\lstset{defaultdialect=[5.3]Lua}
\lstset{defaultdialect=[x86masm]Assembler}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{helvet}
\usepackage{lineno}
\usepackage[affil-it,blocks]{authblk}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage[nomarkers,figuresonly]{endfloat}
%set a box to put the ORCID logo
\newbox{\myorcidaffilbox}
\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{orcid}}

@@

%line spacing
\doublespacing
\usepackage{placeins}
\usepackage{subfig}
\usepackage{breqn}
\usepackage[font={small}]{caption}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

@@

longitudinal data; biomedical data; generalized additive models; simulation; R

@@

Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study). Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others. Tumor response,\textsuperscript{\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}} antibody expression,\textsuperscript{\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}} and cell metabolism\textsuperscript{\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}} are examples of the different situations where researchers have used longitudinal designs to study some physiological response. Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals\textsuperscript{\protect\hyperlink{ref-greening2018}{9}}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients,\textsuperscript{\protect\hyperlink{ref-sio2016}{10}} to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients.\textsuperscript{\protect\hyperlink{ref-kamstra2015}{11}}

Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study. The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times,\textsuperscript{\protect\hyperlink{ref-wagenmakers2008}{12}} and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA). The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}).\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}}

The expected linear behavior of the response through time is a key requisite in rm-ANOVA.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference. In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies. A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings.\textsuperscript{\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}} These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation. Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.

A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data. In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}}. For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid.\textsuperscript{\protect\hyperlink{ref-wang2019}{18}}

Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)),\textsuperscript{\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}} consequently biasing the conclusions of the study. Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni),\textsuperscript{\protect\hyperlink{ref-abdi2010}{21}} but they in turn reduce statistical power (1-\(\beta\))\textsuperscript{\protect\hyperlink{ref-nakagawa2004}{22}}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false).\textsuperscript{\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}} Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.

On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases.\textsuperscript{\protect\hyperlink{ref-ugrinowitsch2004}{25}} Corrections can be made in rm-ANOVA in the absence of compound symmetry,\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}} but the effectiveness of the correction is limited by the size of the sample, the number of measurements\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}, and group sizes.\textsuperscript{\protect\hyperlink{ref-keselman2001}{29}} In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons\textsuperscript{\protect\hyperlink{ref-charan2013}{30}} which might cause the corrections for lack of compound symmetry to be ineffective.

Due to a variety of causes, the number of observations during a study can vary between all subjects. For example, in a clinical trial patients may voluntarily withdraw, whereas attrition due to injury or weight loss in preclinical animal studies is possible. It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points. In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated. When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis.\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13}} This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects. At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data. The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data.\textsuperscript{\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}} Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age). When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent.\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}} And even more importantly, LMEMs also assume a linear relationship between the response and time,\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} making them unsuitable to analyze non-linear data.

As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference. In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data. Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data. Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data,\textsuperscript{\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}} health-environment interactions\textsuperscript{\protect\hyperlink{ref-yang2012}{35}} and the dynamics of government in political science.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and 3) can easily accommodate missing observations. Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.

The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs. In particular, \(\textsf{R}\)\textsuperscript{\protect\hyperlink{ref-r}{38}} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv}\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user. At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models.\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}} Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.

This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas. First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations are explained in detail. Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations. Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies.\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}} The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies. Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in \(\textsf{R}\), in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.

@@

A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}} A LMEM with interaction between time and treatment for a longitudinal study can be written as:

@@

{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_No_Appendix_files/figure-latex/l-q-response-1}

@@

{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_No_Appendix_files/figure-latex/basis-plot-1}
